A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-06 tex/sec3-defs.tex
[ tex ] Use `\aritysymbol` in S3; explain the notation more carefully

\section{Simple Type Theories and Bidirectional Type Systems}\label{sec:defs}
This section provides generic definitions
%\footnote{%
%Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
of simple types, simple type theories, and bidirectional type systems using simply typed $\lambda$-calculus in \Cref{sec:key-ideas} as our running example.

@@

\begin{definition} \label{def:simple-signature}
  \begin{figure}
%    \begin{minipage}[b]{.6\textwidth}
      \centering
      \small
      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Xi$}
      \begin{mathpar}
        \inferrule{\Xi \ni X_i}{\Xi |-_\Sigma X_i} \and
        \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
      \end{mathpar}
      \caption{Simple types}
      \label{fig:simple-type}
%    \end{minipage}
%    \begin{minipage}[b]{.35\textwidth}
%      \centering
%      \small
%      \judgbox{\Xi |-_{\Sigma} \Gamma}{}
%      \begin{mathpar}
%        \inferrule{ }{\Xi |-_{\Sigma} \cdot }\and
%        \inferrule{\Xi |-_{\Sigma} A \\ \Xi|-_{\Sigma} \Gamma}{\Xi |-_{\Sigma} \Gamma, x : A}
%      \end{mathpar}
%      \caption{Contexts}
%    \label{fig:simple-context}
%    \end{minipage}
  \end{figure}
  A \emph{signature} $\Sigma$ for simple types consists of a set\footnote{%
    Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
    As we work with Martin-L\"of type theory with \AxiomK, all types are actually sets in the sense of homotopy type theory where all inhabitants of its identity type are proved equal.
  }
  $I$ with a decidable equality and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
  For a signature $\Sigma$, a \emph{type} $A : \Type_{\Sigma}(\Xi)$ over a variable set $\Xi$ is either
  \begin{enumerate}
    \item a variable in $\Xi$ or
    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$.
  \end{enumerate}
  %as defined in \Cref{fig:simple-type}.
  A \emph{context} $\Gamma \colon \Cxt_{\Sigma}(\Xi)$ over $\Xi$ is a list $\Gamma$ of types over $\Xi$. 
\end{definition}
Note that the second clause is a rule schema giving rise to as many rules as inhabitants $i$ in $I$.

@@

Without the loss of generality, we may use the family of sets $\Fin(n)$ of natural numbers less than $n$ as sets of type variables in this paper, following the inductive definition by \citet{Dybjer1994}.
In bidirectional type synthesis, it is necessary to have a decidable equality on $\Type_{\Sigma}(\Xi)$, as we need to compare the synthesized type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
If $\Xi$ such as $\Fin(n)$ has a decidable equality , the decidable equality on $\Type_{\Sigma}(\Xi)$ can be given.
In addition, the arity of a construct is always finite, so it suffices to use natural numbers to denote type variables in a $\Sigma$-type.
Formally, we restrict the set $\Xi$ of type variables to the family of $\Fin(n)$, but for illustrative purposes we still use named variables for presentation.

As \Cref{def:simple-signature} includes type variables, we introduce the typical meta-operations: substitution.
\begin{definition}
  The \emph{(simultaneous) type substitution} for a function $\rho\colon\Xi \to \Type_{\Sigma}(\Xi')$ is a map, which sends $A : \Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho} : \Type_{\Sigma}(\Xi')$, defined as usual.
  By abuse of notation, the substitution $\simsub{\Gamma}{\rho}$ of a context $\Gamma$ is defined by applying substitution $\simsub{A}{\rho}$ to each type $A$ in the context.
\end{definition}

\subsection{Binding Signatures and Simple Type Theories} \label{subsec:binding-sig}

A simple type theory consists of
\begin{enumerate*}
  \item a set of raw terms $\isTerm{t}$ indexed by a set $V$ of untyped variables where each construct is allowed to bind some variables (as $\Rule{Abs}$) and to take multiple arguments (as $\Rule{App}$);
  \item a set of typing derivations indexed by a list $\Gamma$ of typed variables, a type~$A$, and a raw term $\isTerm{t}$ to set their type constraints. 
\end{enumerate*}
Therefore, to specify a construct like $\Rule{Abs}$ or $\Rule{App}$, we need to enrich the notion of signature with a type and an extension context for variable binding for each argument and a type for the operation:
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
\end{definition}
Each inhabitant $o: O$, meant to represent a term construct $\tmOp_o$ in a simple type theory, has a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A_0\right)$ as its arity $\arity_o$ where
\begin{enumerate}
  \item $\Xi$ is the number of type variables, 
  \item $A : \Type_\Sigma(\Xi)$ is for the type of $\tmOp_o$, and
  \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
    \begin{enumerate}
      \item $\Delta_i : \Cxt_{\Sigma}(\Xi)$ is for the types of bound variables and
      \item $A_i : \Term_{\Sigma}(\Xi)$ for the type of the $i$-th argument of $\tmOp_{o}$.
    \end{enumerate}
\end{enumerate}
For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A_0$ to indicate an operation with its arity and omit $[\Delta]$ if $\Delta$ is empty. 
For example, the $\Rule{Abs}$ rule can be specified by the operation 
\[
  \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B)
\]
indicating that the $\Rule{Abs}$ rule has two type variables $A$ and $B$, one argument of type $B$ and which binds a variable of type $A$, and the type $A \bto B$ for the construct $\lambda$.
Similarly, the $\Rule{App}$ rule can be specified by $\mathsf{app}\colon A, B \rhd A \bto B, A \to B$ where extension contexts for its two arguments are empty.

Also for brevity, a simple type theory $(\Sigma, \Omega)$ refers to the simple type theory specified by some type signature $\Sigma$ and a binding signature $\Omega$.

\begin{definition}
\begin{figure}
  \centering
  \small
  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}}{Given a list V of variables, $\isTerm{t}$ is a raw term for the signature $(\Sigma, \Omega)$ with free variables in $V$}
  \begin{mathpar}
    \inferrule{\isTerm{x} \in V}{V |-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}}{V |-_{\Sigma, \Omega} \isTerm{t \annotate A}}\,\Rule{Anno}
    \and
    \inferrule{V, \vars{x}_\isTerm{1} |-_{\Sigma, \Omega} \isTerm{t_1} \\ \cdots \\ V, \vars{x}_\isTerm{n} |-_{\Sigma, \Omega} \isTerm{t_n}}
    {V |-_{\Sigma, \Omega} \tmOpts }\,\Rule{Op}
  \end{mathpar}
%  \begin{flushright}
%    for $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A$ in $\Omega$
%  \end{flushright}%
  \caption{Raw terms}
  \label{fig:raw-terms}
\end{figure}
  For a simple type theory $(\Sigma, \Omega)$, the set of \emph{raw terms} indexed by a context~$V$ of free variables consists of
  \begin{enumerate}
    \item variables in $V$,
    \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ in $V$ and a type $A$, and
    \item a construct $\tmOpts$ for some $o \colon \Xi \rhd [\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n} \to A_0$ in $O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$ and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$
  \end{enumerate}
  corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
\end{definition}

The definition of typing derivations is slightly involved.
We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
Hence we choose to include a substitution as part of its typing derivation explicitly:
\begin{definition}\label{def:typing-derivations}
  \begin{figure}
    \centering
    \small
    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $\isTerm{t}$ has a type $A$ without type variables under $\Gamma$ for the signature $(\Sigma, \Omega)$}
    \begin{mathpar}
      \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
      \and
      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) : A}\,\Rule{Anno}
      \and
      \inferrule{\rho : \Sub{\Xi}{\emptyset} \\  \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
      {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}}\,\Rule{Op}
    \end{mathpar}
    \caption{Typing derivations}
    \label{fig:extrinsic-typing}
  \end{figure}
  For a simple type theory $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ indexed by a context $\Gamma$, a raw term with free variables in $\erase{\Gamma}$, and a type $A$ consists of 
  \begin{enumerate}
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} x : A$ if $x : A$ is in $\Gamma$,
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotate A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $o \colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A_0$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
  \end{enumerate}
  corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing} respectively.
\end{definition}
Note that, by definition, our simple type theory is always syntax-directed, since each rule in \Cref{fig:raw-terms} corresponds exactly to one rule in \Cref{fig:extrinsic-typing}. 
The $\Rule{Op}$ in both figures is a rule schema and gives rise to as many rules as there are operations $o : O$.

\begin{example}
  Raw terms (\Cref{fig:STLC-raw-terms}) and typing derivations (\Cref{fig:STLC-typing-derivations}) for simply typed $\lambda$-calculus can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the binding signature consisting of 
 \[
   \mathsf{app}\colon A, B \rhd A \bto B, A \to B
   \quad\text{and}\quad
   \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B).
 \]
 Rules $\Rule{Abs}$ and $\Rule{App}$ in simply typed $\lambda$-calculus are subsumed by the $\Rule{Op}$ rule schema, as an application $t\;u$ and an abstraction $\lam{x}t$ can be introduced as $\tmOp_{\mathsf{app}}(t, u)$ and $\tmOp_{\mathsf{abs}}(x.t)$, respectively.
\end{example}

\subsection{Bidirectional Binding Signatures and Bidirectional Type Systems} \label{subsec:bidirectional-system}
Finally, for a bidirectional type system, we have raw terms and typing derivations like a simple type theory, but its typing judgements come in two forms $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$.
Observe that these two typing judgements are just one typing judgement $\Gamma |- t :^\dir{d} A$ indexed additionally by a \emph{mode} $d : \Mode$---which is either $\syn$ or $\chk$---so that $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$ are just abbreviations of $\Gamma |- t :^{\syn} A$ and $\Gamma |- t :^{\chk} A$ respectively.
Hence, to specify a bidirectional type system, we need further enrich the notion of binding signature with modes as follows.

@@

For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n} \to A_0^{\dir{d}}$ to indicate an operation with its arity. 
For example, the $\ChkRule{Abs}$ rule in \Cref{fig:STLC-bidirectional-typing-derivations} can be specified by $\mathsf{abs}\colon A , B \rhd [A]B^{\chk} \to (A \bto B)^{\chk}$ indicating additionally that the construct $\lambda$ and its argument are in checking mode.
Similarly, the $\SynRule{App}$ rule can be specified by $\mathsf{app}\colon A, B \rhd (A \bto B)^{\syn}, A^{\chk} \to B^{\syn}$.

We also say a bidirectional type system $(\Sigma, \Omega)$ referring the bidirectional type system specified by some type signature $\Sigma$ and a bidirectional binding signature $\Omega$.

\begin{remark}
  Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, so a bidirectional binding signature $\Omega$ also specifies a simple type theory, including its raw terms and typing derivations.
  We call $\erase{\Omega}$ the \emph{(mode) erasure} of $\Omega$.
\end{remark}

>>>>

\section{Simple Type Systems and Bidirectional Type Systems}\label{sec:defs}
This section provides generic definitions
%\footnote{%
%Here's the 'small print': Our definitions cover typical examples and set the scope for bidirectional simple type systems discussed, but do not offer comprehensive coverage of all possibilities.}
of simple types, simple type systems, and bidirectional type systems using simply typed $\lambda$-calculus in \Cref{sec:key-ideas} as our running example.

@@

\begin{definition} \label{def:simple-signature}
%  \begin{figure}
%%    \begin{minipage}[b]{.6\textwidth}
%      \centering
%      \small
%      \judgbox{\Xi|-_{\Sigma} A}{$A$ is a well-formed type with type variables in $\Xi$}
%      \begin{mathpar}
%        \inferrule{\Xi \ni X_i}{\Xi |-_\Sigma X_i} \and
%        \inferrule{\Xi |-_{\Sigma} A_1 \\ \cdots \\ \Xi |-_{\Sigma} A_n}{\Xi |-_{\Sigma} \tyOp_i(A_1, \ldots, A_n)}\;\text{where $n = \arity(i)$}
%      \end{mathpar}
%      \caption{Simple types}
%      \label{fig:simple-type}
%%    \end{minipage}
%%    \begin{minipage}[b]{.35\textwidth}
%%      \centering
%%      \small
%%      \judgbox{\Xi |-_{\Sigma} \Gamma}{}
%%      \begin{mathpar}
%%        \inferrule{ }{\Xi |-_{\Sigma} \cdot }\and
%%        \inferrule{\Xi |-_{\Sigma} A \\ \Xi|-_{\Sigma} \Gamma}{\Xi |-_{\Sigma} \Gamma, x : A}
%%      \end{mathpar}
%%      \caption{Contexts}
%%    \label{fig:simple-context}
%%    \end{minipage}
%  \end{figure}
  A \emph{signature} $\Sigma$ for simple types consists of a set\footnote{%
    Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
    As we work in Martin-L\"of type theory with \AxiomK, all types are legitimately sets in the sense of homotopy type theory where all inhabitants of its identity type are proved equal.
  }
  $I$ with a decidable equality and an \emph{arity} function $\arity\colon I \to \mathbb{N}$.
  For a signature $\Sigma$, a \emph{type} $A : \Type_{\Sigma}(\Xi)$ over a variable set $\Xi$ is either
  \begin{enumerate}
    \item a variable in $\Xi$ or
    \item $\tyOp_{i}(A_1, \ldots, A_n)$ for some $i:I$ with $\arity(i) = n$ and types $A_1,\ldots, A_n$.
  \end{enumerate}
  %as defined in \Cref{fig:simple-type}.
  A \emph{context} $\Gamma \colon \Cxt_{\Sigma}(\Xi)$ over $\Xi$ is a list $\Gamma$ of types over $\Xi$. 
\end{definition}

@@

As \Cref{def:simple-signature} includes type variables, we introduce the typical meta-operations: substitution.
\begin{definition}
  The \emph{(simultaneous) type substitution} for a function $\rho\colon\Xi \to \Type_{\Sigma}(\Xi')$ is a map which sends $A : \Type_{\Sigma}(\Xi)$ to $\simsub{A}{\rho} : \Type_{\Sigma}(\Xi')$ and is defined as usual.
  By abuse of notation, the substitution $\simsub{\Gamma}{\rho}$ of a context $\Gamma$ is defined by applying substitution $\simsub{A}{\rho}$ to each type $A$ in the context.
\end{definition}

We may confine arbitrary variable set $\Xi$ to the family of sets $\Fin(n)$ of natural numbers less than $n$ as sets of type variables, following the inductive definition by \citet{Dybjer1994}, for the reasons below.
In bidirectional type synthesis, it is necessary to have a decidable equality on $\Type_{\Sigma}(\Xi)$, as we need to compare the synthesized type $B$ with a given type $A$ for the $\Rule{Sub}$ rule.
If $\Xi$ such as $\Fin(n)$ has a decidable equality, the decidable equality on $\Type_{\Sigma}(\Xi)$ can be given.
In addition, the arity of a construct is always finite, so it suffices to use natural numbers to denote type variables in a type.
Therefore, we formally restrict the set $\Xi$ of type variables to the family of $\Fin(n)$, but for illustrative purposes we still use named variables for presentation.
\subsection{Binding Signatures and Simple Type Systems} \label{subsec:binding-sig}

A simple type system consists of
\begin{enumerate*}
  \item a set of raw terms $\isTerm{t}$ indexed by a set $V$ of untyped variables where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
  \item a set of typing derivations indexed by a list $\Gamma$ of typed variables, a type~$A$, and a raw term $\isTerm{t}$ to set their type constraints. 
\end{enumerate*}
Therefore, to specify a construct like $\Rule{Abs}$ and $\Rule{App}$, we need to enrich the notion of signature with a type and an extension context for variable binding for each argument and a type for the operation:
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \mathit{ar}\colon O \to \sm{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
\end{definition}
Each inhabitant $o: O$, meant to represent a term construct $\tmOp_o$ in a simple type system, has a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A_0\right)$ as its arity $\arity_o$ where
\begin{enumerate}
  \item $\Xi$ is the number of type variables that are allowed to appear in a typing rule, 
  \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs for specifying arguments where
    \begin{enumerate}
      \item $\Delta_i : \Cxt_{\Sigma}(\Xi)$ is the \emph{extension context} consisting of types for bound variables and
      \item $A_i : \Term_{\Sigma}(\Xi)$ is for the type of the $i$-th argument (i.e.\ premise) of $\tmOp_{o}$;
    \end{enumerate}
  \item $A_0 : \Type_\Sigma(\Xi)$ is for the type of the construct $\tmOp_o$ in the conclusion.
\end{enumerate}
For brevity, we write $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}, \ldots, \left[\Delta_{n}\right] A_{n}}{A_0}$ to indicate an operation with its arity and omit $[\Delta]$ if $\Delta$ is empty. 
For example, the $\Rule{Abs}$ rule in \Cref{fig:STLC-typing-derivations} can be specified by the operation 
\[
  \aritysymbol{\mathsf{abs}}{(A, B)}{[A]B}{(A \bto B)}
\]
indicating that in the $\Rule{Abs}$ rule for a raw term $\lam{x}t$
\begin{enumerate*}
  \item there are \emph{two} type variables named $A$ and $B$;
  \item an argument (i.e.\ a typing derivation of $\Gamma, x : A |- t : B$) of type $B$ in a context extended by a variable $x$ of type $A$;
  \item the type $A\bto B$ for the conclusion.
\end{enumerate*}
Similarly, the $\Rule{App}$ rule for $t\;u$ can be specified by $\aritysymbol{\mathsf{app}}{(A, B)}{(A \bto B), A}{B}$ which takes two arguments (i.e.\ typing derivations for $t$ and $u$) under the same context so that their extension contexts are both empty.

Also for brevity, a simple type system $(\Sigma, \Omega)$ refers to the simple type system specified by some type signature $\Sigma$ and a binding signature $\Omega$.

\begin{definition}
\begin{figure}
  \centering
  \small
  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}}{Given a list V of variables, $\isTerm{t}$ is a raw term for the signature $(\Sigma, \Omega)$ with free variables in $V$}
  \begin{mathpar}
    \inferrule{\isTerm{x} \in V}{V |-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}}{V |-_{\Sigma, \Omega} \isTerm{t \annotate A}}\,\Rule{Anno}
    \\
    \inferrule{V, \vars{x}_\isTerm{1} |-_{\Sigma, \Omega} \isTerm{t_1} \\ \cdots \\ V, \vars{x}_\isTerm{n} |-_{\Sigma, \Omega} \isTerm{t_n} } {V |-_{\Sigma, \Omega} \tmOpts }\,\Rule{Op} \and
    \text{for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n}}{A_0}$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms}
  \label{fig:raw-terms}
\end{figure}
  For a simple type system $(\Sigma, \Omega)$, the set of \emph{raw terms} indexed by a context~$V$ of free variables consists of
  \begin{enumerate}
    \item variables in $V$,
    \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ in $V$ and a type $A$, and
    \item a construct $\tmOpts$ for some $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n}}{A_0}$ in $O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$ and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$
  \end{enumerate}
  corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
\end{definition}

The definition of typing derivations is a bit more involved.
We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
Hence we choose to include a substitution as part of its typing derivation explicitly:
\begin{definition}\label{def:typing-derivations}
  \begin{figure}
    \centering
    \small
    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{A raw term $\isTerm{t}$ has a type $A$ without type variables under $\Gamma$ for the signature $(\Sigma, \Omega)$}
    \begin{mathpar}
      \inferrule{(\isTerm{x} : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} : A}\,\Rule{Var}
      \and
      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) : A}\,\Rule{Anno}
      \and
      \inferrule{\rho : \Sub{\Xi}{\emptyset} \\  \Gamma, \vec{\isTerm{x}}_\isTerm{1} : \simsub{\Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \vec{\isTerm{x}}_\isTerm{n} : \simsub{\Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
      {\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}}\,\Rule{Op}
      \\
    \text{for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}, \ldots, [\Delta_{n}] A_{n}}{A_0}$ in $\Omega$}
    \end{mathpar}
    \caption{Typing derivations}
    \label{fig:extrinsic-typing}
  \end{figure}
  For a simple type system $(\Sigma, \Omega)$, the set of \emph{typing derivations} of $\Gamma \vdash \isTerm{t} : A$ indexed by a context $\Gamma$, a raw term with free variables in $\erase{\Gamma}$, and a type $A$ consists of 
  \begin{enumerate}
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} x : A$ if $x : A$ is in $\Gamma$,
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} (t \annotate A) : A$ if $\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation, and
    \item a derivation of $\Gamma |-_{\Sigma, \Omega} \tmOpts : \simsub{A_0}{\rho}$ for $\aritysymbol{o}{\Xi}{[\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n}}{A_0}$ if there is a function $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \Delta_i \vdash_{\Sigma, \Omega} \isTerm{t_i} : \simsub{A_i}{\rho}$ for each $i$,
  \end{enumerate}
  corresponding to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:extrinsic-typing} respectively.
\end{definition}
Note that our simple type system is always syntax-directed, since each rule in \Cref{fig:raw-terms} corresponds exactly to one rule in \Cref{fig:extrinsic-typing}. 
The $\Rule{Op}$ rule in both figures is a rule schema and gives rise to as many rules as there are operations $o : O$.

\begin{example}
  Raw terms (\Cref{fig:STLC-raw-terms}) and typing derivations (\Cref{fig:STLC-typing-derivations}) for simply typed $\lambda$-calculus can be specified by the type signature $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) and the binding signature consisting of 
 \[
   \aritysymbol{\mathsf{app}}{(A, B)}{(A \bto B), A}{B}
   \quad\text{and}\quad
   \aritysymbol{\mathsf{abs}}{(A , B)}{[A]B}{(A \bto B)}.
 \]
 Rules $\Rule{Abs}$ and $\Rule{App}$ in simply typed $\lambda$-calculus are subsumed by the $\Rule{Op}$ rule schema, as an application $t\;u$ and an abstraction $\lam{x}t$ can be introduced as $\tmOp_{\mathsf{app}}(t, u)$ and $\tmOp_{\mathsf{abs}}(x.t)$, respectively.
\end{example}

\subsection{Bidirectional Binding Signatures and Bidirectional Type Systems} \label{subsec:bidirectional-system}
Finally, for a bidirectional type system, we have raw terms and typing derivations like a simple type system, but its typing judgements come in two forms $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$.
Observe that these two typing judgements are just one typing judgement $\Gamma |- t :^\dir{d} A$ indexed additionally by a \emph{mode} $d : \Mode$---which is either $\syn$ or $\chk$---so that $\Gamma |- t \syn A$ and $\Gamma |- t \chk A$ are just abbreviations of $\Gamma |- t :^{\syn} A$ and $\Gamma |- t :^{\chk} A$ respectively.
Hence, to specify a bidirectional type system, we need further enrich the notion of binding signature with modes as follows.

@@

For brevity, we write $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A^{\dir{d_n}}_{n}}{A_0^{\dir{d}}}$ to indicate an operation with its arity. 
For example, the $\ChkRule{Abs}$ rule in \Cref{fig:STLC-bidirectional-typing-derivations} can be specified by $\aritysymbol{\mathsf{abs}}{A , B}{[A]B^{\chk}}{(A \bto B)^{\chk}}$ indicating additionally that $\lam{x}t$ and its argument $t$ are in checking mode.
Similarly, the $\SynRule{App}$ rule can be specified by $\aritysymbol{\mathsf{app}}{A, B}{(A \bto B)^{\syn}, A^{\chk}}{B^{\syn}}$.

We also say a bidirectional type system $(\Sigma, \Omega)$ referring to the bidirectional type system specified by some type signature $\Sigma$ and a bidirectional binding signature $\Omega$.

\begin{remark}
  Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, so a bidirectional binding signature $\Omega$ also specifies a simple type system, including its raw terms and typing derivations.
  We call $\erase{\Omega}$ the \emph{(mode) erasure} of $\Omega$.
\end{remark}

We generalise bidirectional typing rules (\Cref{fig:STLC-bidirectional-typing-derivations}) as well as mode derivations (\Cref{fig:STLC-mode-derivations}) for simply typed $\lambda$-calculus to derivations for a bidirectional type system $(\Sigma, \Omega)$ where $\Omega$ is some bidirectional binding signature, similar to \Cref{def:typing-derivations} but with an additional $\ChkRule{Sub}$ rule:
\begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
  \begin{figure}
    \centering
    \small
    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{A raw term $\isTerm{t}$ is of type $A$ without type variables in mode $\dir{d}$ under $\Gamma$}
    \begin{mathpar}
      \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
      \and
      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) :^{\syn}  A}\,\SynRule{Anno}
      \and
      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ B = A}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
      \\
      \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\vars{x}_\isTerm{1} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
        \cdots \\
      \Gamma, \simsub{\vars{x}_\isTerm{n} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
      {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}} \,\mathsf{Op}
      \\ \text{for $\aritysymbol{o}{\Xi}{[\Delta_1]A_{1}^{\dir{d_1}}, \ldots, [\Delta_{n}] A_{n}^{\dir{d_n}}}{A_0^{\dir{d}}}$ in $\Omega$}
    \end{mathpar}
    Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$
    \caption{Bidirectional typing derivations}
    \label{fig:bidirectional-typing-derivations}
  \end{figure}
  \begin{figure}
    \centering
    \small
    \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{A raw term $\isTerm{t}$ is in mode $d$ with free variables in $V$}
    \begin{mathpar}
      \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
      \and
